The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-21 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Work on article

\section{Data description}

@@

We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The observation nights are listed in Table~\ref{tbl:obs-list}.

@@

\section{Detection results}
\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observations except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and because of RFI leakage, but because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. Other than that, the RFI leakage is minimal. With exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from Digital Video Broadcasting (DVB), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DVB bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. When observing at 182~MHz, this type of RFI makes most of the 30.72~MHz bandwidth unusable when it is visible. }
\label{fig:dvb}
\end{figure*}

Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive for steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DVB interference from the data statistics. An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting. These statistics are collected by \textsc{cotter}. When DVB is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms, such as having to account for lower SNR and a decreased $uv$ coverage, in place to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%. This is detailed in the last column of Table~\ref{tbl:obs-list}.

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. Detection of this kind of RFI requires flagging at high time and frequency resolution.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged}
\label{fig:150_2}
}\\\vspace{3mm}%
\subcaptionbox{A weakly-observed transmitter centred on 156.66 MHz. This same frequency is flagged in several correlations, and it can therefore be assumed this is real RFI. This is an example in which the \textsc{aoflagger} detects an event that would not be detected by regular thresholding. }[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}
\label{fig:156_7}
}%
\caption{Examples of RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontally flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\end{figure*}

\noindent\begin{figure*}%
% obs data: 1075641128
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-noflags}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-withflags}
\end{center}
\caption{???. }
\label{fig:dvb-burst}
\end{figure*}

%\noindent\begin{figure*}
%\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-eor}\vspace{-1cm}
%\caption{..}
%\label{fig:rfi-per-set-eor}
%\end{center}
%\end{figure*}

%\noindent\begin{figure*}
%\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/plot-rfi-per-set-gleam}\vspace{-1cm}
%\caption{..}
%\label{fig:rfi-per-set-gleam}
%\end{center}
%\end{figure*}

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
\caption{Log N log S}
\label{fig:lognlogs}
\end{center}
\end{figure*}

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases in total the runtime by approximately a percent. On a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

\begin{itemize}
 \item Describe RFI examples (EoR low band \& high band)
 \item Log N log S Histograms
\end{itemize}

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
\end{itemize}
\subsection{Comparison with LOFAR}

\section{Conclusions}
\begin{itemize}
 \item Much less RFI than LOFAR (but at a cost)
 \item About half the observations don't need flagging at all. 
 \item FM not completely empty
\end{itemize}

>>>>

@@

\section{Data description}
We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The observation nights are listed in Table~\ref{tbl:obs-list}.

@@

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/rfi-per-band}
\caption{RFI occupancy per subband, calculated over all observations except GLEAM 2014-03-17. The latter has been left out because it is affected by DVB.}
\label{fig:rfi-per-band}
\end{center}
\end{figure*}

\section{Detection results}
The fifth column of Table~\ref{tbl:obs-list} lists for each of the observation nights the ratio of samples that were classified as RFI by the \textsc{aoflagger}. This includes false detections, which are estimated in \citet{lofar-radio-environment} to account for approximately 0.4\% detections. Compared to the EoR observations, the GLEAM observations have a higher average of 1.05\% of RFI. This is mostly caused by the FM bands, which are only observed in GLEAM observations. The EoR low-band night shows 0.81\% RFI, while the EoR high-band observations have an average of 0.58\%. The total RFI occupancy in all observations is 0.96\%. Weighting the occupancy in each channel by the inverse time that it is observed results in an occupancy of 1.13\%, which is the global expected RFI ratio in the 72.3--230.8 MHz range.

@@

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and because of RFI leakage, but because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}\hspace*{-0.2cm}\includegraphics[width=10cm]{img/dvb_example}\includegraphics[width=5cm]{img/2014-03-17-dvb-stddevs}
\end{center}
\caption{RFI from Digital Video Broadcasting (DVB), which is visible in 1 of the 9 observations (GLEAM~2014-03-17) that cover the DVB frequencies. Left plot: dynamic spectrum for the worst affected snapshot; right plot: visibility RMS in a few of the DVB bands for the affected night after flagging. Because of its broadband nature, this kind of RFI is not well flagged in the initial \textsc{aoflagger} step, but it is detectable in the global statistics. }
\label{fig:dvb}
\end{figure*}

Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive for steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

Although the flagger does not adequately flag DVB, the right plot of Fig.~\ref{fig:dvb} shows that it is possible to determine whether a snapshot is affected by DVB interference from the data statistics. An option is to test whether the visibility RMS of a snapshot in any of the occupied DVB bands exceeds the RMS in DVB band 4, which is not used for broadcasting. These statistics are calculated by \textsc{cotter}, hence this can be validated without having to read the data again. When DVB is detected in a snapshot, the entire snapshot can be removed or the residual unaffected data within the snapshot can be used. However, it might not be worth it to have special mechanisms in place, such as having to account for lower SNR and a decreased $uv$ coverage, to salvage the residual good data within an affected snapshot. Because 2 out of 7 hours are affected in 1 out of the 9 randomly selected nights, the probability of observing DVB is roughly 3\%. This is detailed in the last column of Table~\ref{tbl:obs-list}.

\noindent\begin{figure*}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=18cm]{img/lognlogs-multiplot}
\caption{Visibility amplitude distributions with logarithmic axes. The dashed line represent fits over a reasonably constant part of the tail (selected by eye) of the RFI distribution: $N=\beta S^{-1.43}$ for GLEAM, $N=\beta S^{-4.47}$ for EoR high and $N=\beta S^{-1.33}$ for EoR low.}
\label{fig:lognlogs}
\end{center}
\end{figure*}

\subsection{Distribution analysis}
In \citet{offringa-rfi-distributions}, using LOFAR it is shown that when many interfering transmitters are observed, the distribution of the visibility brightness will have a power-law tail with an exponent of approximately -1.5 to -1.6. For the GLEAM, EoR high-band and EoR low-band observations of this work, the distributions are drawn in Fig.~\ref{fig:lognlogs}. The three distributions are different due to the fact that each observation covers different frequencies.  Separate distributions are shown for visibilities that have been detected as RFI (red lines) and those that have been classified as RFI free (green lines). The EoR high-band curves do not include the night that was contaminated by DVB RFI.

In the ideal case, the residual visibility distributions show a smooth Rayleigh curve (see \citealt{offringa-rfi-distributions}). The tails of the residual curves of Fig.~\ref{fig:lognlogs} are however not smooth. Especially in the case of the EoR high-band and low-band observations, the residual curve shows RFI that has not been properly detected as such. In the case of EoR low-band, samples are not detected as RFI even though they are an order of magnitude brighter than where the GLEAM residual curve ends. The reason such samples leak through detection is because the detection strategy includes a bandpass correction per 1.28-MHz sub-band. When RFI persistently contaminates approximately a full subband, its effect will be absorbed in the bandpass correction and becomes invisible for the flagger. In some cases, this will cause the calibration to fail on those channels, after which it will be flagged nevertheless. In other cases, such RFI is included in the imaging step or further processing, which is obviously very undesirable.

The distribution curves do not show a very significant power-law tail. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being an arbitrary scaling factor. The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected. Fits over by-eye selected regions have been added, with $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves of Fig.~\ref{fig:lognlogs}, respectively.

\noindent\begin{figure*}
\subcaptionbox{RFI contamination found in the 2 m amateur band (146 MHz).}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/amateur-2m-band-example}\includegraphics[width=9.1cm]{img/amateur-2m-band-flagged}
\label{fig:amateur-2m}
}\\\vspace{3mm}%
\subcaptionbox{Short RFI burst of 2 s centred on 150.17 MHz. Detection of this kind of RFI requires flagging at high time and frequency resolution.}[\linewidth]{%
% obs data: 1061583576, 2013-08-26.
\includegraphics[width=9.1cm]{img/150_2_mhz_example_ann}\includegraphics[width=9.1cm]{img/150_2_mhz_flagged_ann}
\label{fig:150_2}
}%
\caption{RFI events found in the EoR low band (138.9--169.6 MHz) in a 2-min snapshot with relatively high RFI contamination. These panels show the Stokes I amplitudes. In the right figures, the result of RFI detection is shown with purple. The horizontal flagged lines are flagged because they are 1.28-MHz subband edge or centre channels, which are unusable because of aliasing of the poly-phase filter bank and DC offsets, respectively.}
\label{fig:2m-amateur-band}
\end{figure*}

\noindent\begin{figure*}%
% obs data: 1075641128
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-noflags}\includegraphics[width=8cm]{img/EoR-high-band-DVB-burst-example-withflags}
\end{center}
\caption{DVB RFI briefly visible in an EoR high-band observation. The plots show the Stokes~I amplitudes for a single correlation. The right plot shows the result of RFI detection marked in purple.}
\label{fig:dvb-burst}
\end{figure*}

\noindent\begin{figure*}%
% obs data: 1075637224
\begin{center}\hspace*{-0.2cm}\includegraphics[width=8cm]{img/broadband-pulse-unflagged}\includegraphics[width=8cm]{img/broadband-pulse-flagged}
\end{center}
\caption{Broadband pulses found in an EoR high-band observation. The two panels show Stokes~I visibility amplitudes of a single correlation. In the right plot, the RFI detection flags are marked with purple. Each event occupies 2 timesteps of 0.5~s.}
\label{fig:broadband-pulse}
\end{figure*}

\noindent\begin{figure*}%
% obs data: 1061583576, 2013-08-26.
\begin{center}
\includegraphics[width=9.1cm]{img/156_7_mhz_example}\includegraphics[width=9.1cm]{img/156_7_mhz_flagged}\\
\includegraphics[width=9.1cm]{img/156_7_mhz_stddev}\includegraphics[width=9.1cm]{img/156_7_mhz_stddev_flagged}
\caption{A weakly-observed transmitter centred on 156.66 MHz. Top-left window: a single correlation, where the transmitter is barely visible; Top-right window: detected RFI marked with purple. When performing per-baseline flagging, the transmitter is partly detected; Bottom-left window: when combining data from all baselines, the transmitter becomes clearly visible; Bottom-right window: in yellow, result of executing \textsc{aoflagger} on the standard deviations calculated over all baselines.}\label{fig:156_7}
\end{center}
\end{figure*}

\subsection{RFI types}
A variety of RFI events types are observed in the data sets. While there are too many transmitters to show examples for each, it is helpful to understand what kind of events are visible and how they are flagged by the \textsc{aoflagger}. Therefore, a few typical examples are shown.

Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz). This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all. This variability might be caused by intrinsic variation or movement of the transmitter or varying propagation conditions. While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot. In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is showed but is zoomed in on a briefly-observed RFI event at 150.17 MHz. This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; TODO cite) might help to analyze these phenomena.

Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation. The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter is not completely detected (Fig.~\ref{fig:156_7}, bottom-left plot). The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot). This kind of detection is currently not implemented in \textsc{cotter}.

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. On a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
 \item The presence of strong interference implies there might be weak interference present that is easily seen or detected. 
\end{itemize}
\subsection{Comparison with LOFAR}

\section{Conclusions}
\begin{itemize}
 \item Much less RFI than LOFAR (but at a cost)
 \item About half the observations don't need flagging at all. 
 \item FM not completely empty
 \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
\end{itemize}
